---
title: "Course Project 2- Stormdata.RMD"
output: html_document
---
#Synopsys 
This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

##Assignment 
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. We must use the database to answer the questions below and show the code for our entire analysis. Our analysis can consist of tables, figures, or other summaries. We may use any R package we want to support our analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Data Processing
```{r}
#Downloading file
if(!file.exists("stormData.csv.bz2")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
  destfile = "stormData.csv.bz2", method = "curl")
}
```

```{r}
#loading data
NOAAStormdata<-read.csv("stormData.csv.bz2")
```

subset the required columns 
```{r}
#colnames(NOAAStormdata)
#subsetted colums indicate the effects of event-types
ReqNOAA<-NOAAStormdata[,c('EVTYPE','FATALITIES','INJURIES', 'PROPDMG', 'PROPDMGEXP', 'CROPDMG', 'CROPDMGEXP')]
```


```{r}
#Converting H K M B of PROPDMGEXP to units
ReqNOAA$PROPDMGNUM=0
ReqNOAA[ReqNOAA$PROPDMGEXP=="H", ]$PROPDMGNUM = ReqNOAA[ReqNOAA$PROPDMGEXP == "H", ]$PROPDMG * 10^2
ReqNOAA[ReqNOAA$PROPDMGEXP=="K", ]$PROPDMGNUM = ReqNOAA[ReqNOAA$PROPDMGEXP == "K", ]$PROPDMG * 10^3
ReqNOAA[ReqNOAA$PROPDMGEXP=="M", ]$PROPDMGNUM = ReqNOAA[ReqNOAA$PROPDMGEXP == "M", ]$PROPDMG * 10^6
ReqNOAA[ReqNOAA$PROPDMGEXP=="B", ]$PROPDMGNUM = ReqNOAA[ReqNOAA$PROPDMGEXP == "B", ]$PROPDMG * 10^9
```


```{r}
#Converting H K M B of CROPDMGEXP units
ReqNOAA$CROPDMGNUM=0
ReqNOAA[ReqNOAA$CROPDMGEXP=="H",]$CROPDMGNUM=ReqNOAA[ReqNOAA$CROPDMGEXP=="H",]$CROPDMG * 10^2
ReqNOAA[ReqNOAA$CROPDMGEXP=="K",]$CROPDMGNUM=ReqNOAA[ReqNOAA$CROPDMGEXP=="K",]$CROPDMG * 10^3
ReqNOAA[ReqNOAA$CROPDMGEXP=="M",]$CROPDMGNUM=ReqNOAA[ReqNOAA$CROPDMGEXP=="M",]$CROPDMG * 10^6
ReqNOAA[ReqNOAA$CROPDMGEXP=="B",]$CROPDMGNUM=ReqNOAA[ReqNOAA$CROPDMGEXP=="B",]$CROPDMG * 10^9
```
###Results
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```{r}
#plotting no of fatalities according to Event Types
library(ggplot2)
fatalities<-aggregate(FATALITIES ~ EVTYPE, data=ReqNOAA, sum)
fatalities<-fatalities[order(-fatalities$FATALITIES), ][1:20, ]
fatalities$EVTYPE<-factor(fatalities$EVTYPE, levels = fatalities$EVTYPE)

ggplot(fatalities, aes(x = EVTYPE, y = FATALITIES)) + 
    geom_bar(stat = "identity", fill="salmon") + theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
    xlab("Event Type") + ylab("Fatalities") + ggtitle("No of fatalities by top 20 Weather Events")
```

```{r}
library(ggplot2)
#aggregating injuries
injuries <- aggregate(INJURIES ~ EVTYPE, data=ReqNOAA, sum)
injuries <- injuries[order(-injuries$INJURIES), ][1:20, ]
injuries$EVTYPE <- factor(injuries$EVTYPE, levels = injuries$EVTYPE)

##plotting no of injuries according to Event Types
ggplot(injuries, aes(x = EVTYPE, y = INJURIES)) + geom_bar(stat = "identity", fill="purple") + 
  theme(axis.text.x = element_text(angle = 70, hjust = 1)) + 
    xlab("Event Type") + ylab("Injuries") + ggtitle("No of injuries by top 20 Weather Events")
```

2. Across the United States, which types of events have the greatest economic consequences?

```{r}
library(ggplot2)
#aggregating property and crop damages
damages<-aggregate(PROPDMGNUM + CROPDMGNUM ~ EVTYPE, data=ReqNOAA, sum)
names(damages)=c("EVTYPE", "TOTALDAMAGE")
damages<-damages[order(-damages$TOTALDAMAGE), ][1:20, ]
damages$EVTYPE<-factor(damages$EVTYPE, levels = damages$EVTYPE)

#plot
ggplot(damages, aes(x = EVTYPE, y = TOTALDAMAGE)) + 
    geom_bar(stat = "identity", fill="skyblue") + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    xlab("Event Type") + ylab("Damages in $") + ggtitle("Property & Crop Damages by top 20 Weather Events")
```

